---
title: "Denali_ordination_final"
author: "Mélanie Jean"
date: "23/04/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd("~/Dropbox/PDF NAU/Denali Fire/Submission/Re-analysis")
```


## Redo the analyses for the ordination using modified cover data
Here, we use the cover data measured on the transects before and after fire, and complete it by adding a "presence", i.e. 0.1 value to all the species that were found in the full plot survey data, but not captured by the transects. 

Open useful librairies
```{r results='hide', message=FALSE, warning=FALSE}
library(vegan)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(vegan)
library(scales)
library(reshape2)
library(stringr)
library(stargazer)
library(directlabels)
library(indicspecies)
library(xtable)
library(ggrepel)
library(gtools)
library(cowplot)
library(egg)
```

Import species data. 
```{r import data}
# Load the cover data
cover=read.csv("Denali_cover_April2021.csv", header=TRUE)
names(cover)
head(cover)
str(cover)
dim(cover)
na.omit(cover)

# Load the frequency data that was converted to 0.5 if present
freq=read.csv("Frequency_to_presence.csv", header=TRUE)
names(freq)
str(cover)
head(freq)
dim(freq)
na.omit(freq)

### All columns, nested for loops!
for (k in seq(6,ncol(cover))){
for (i in seq(1,nrow(cover))){
  if(cover[i,k]<=0){
    cover[i,k]=freq[i,k]}
}
}
cover_merged=cover

write.csv(cover_merged, "cover_merged.csv")

cover2=read.csv("cover_merged.csv", header=TRUE)[,-1]
head(cover2)
```


Data was pre removed from WC23 and WC25, which removed a few species from the dataset that were only found there. 
Other species were found at only one site x sampling time and were removed. Some researchers suggest eliminating species which occur at single sites because of the inflated correlations created by attempting to relate potentially random features at that site to its lone occurrence (Legendre and Legendre, 1998; Poos and Jackson 2012). 
```{r}
denali=cover2
names(denali)
denali$PrePost=as.factor(denali$PrePost)
denali$GRID_ID=as.factor(denali$GRID_ID)

# Now, remove all species columns that sum to zero or that are present at only one site. 
denali.spe=denali[,6:ncol(denali)]
dim(denali.spe) # 103 columns

colSums(denali.spe) # poapal and carcan
colSums(denali.spe != 0) # Will have to remove all the columns that have a colSums =<1

denali.spe.red = denali.spe[,colSums(denali.spe != 0) > 1.5]
dim(denali.spe.red) # 80 columns - 23 species dropped

rowSums(denali.spe.red != 0) # Verify that there are no empty sites - ok 

# Add row names
rownames(denali.spe.red)=denali$GridPlotYear
```



Create the Bray-Curtis distance-matrix associated with the species.
```{r distance matrix}
# All sites and grids
spe=denali.spe.red 
total.d=vegdist(spe, method='bray')
```


Include the environmental variables and spruce density. 
```{r}
env.total=read.csv("Site_Denali_Apr2021_2.csv", header=T)
names(env.total)

# All sites and grids
total.table=full_join(denali, env.total)%>%
    filter(!Grid_plot %in% c("WC25", "WC23", "ET26", "ET27", "ET28", "WC26", "WC27", "WC28"))%>%
    dplyr::select(PrePost, GRID_ID, Grid_plot, GridPlotYear, density_spruce) %>%
    plyr::mutate(density_spruce1, group1=quantcut(density_spruce, q=3))%>%
    mutate(Density_code=plyr::revalue(group1, c("0"="low", "(0,0.0249]"="mid", "(0.0249,0.229]"="high")))

total.table$Density_code=as.factor(total.table$Density_code)
total.table$PrePost=as.factor(total.table$PrePost)
total.table$GRID_ID=as.factor(total.table$GRID_ID)
total.table$Grid_plot=as.factor(total.table$Grid_plot)
total.table$GridPlotYear=as.factor(total.table$GridPlotYear)

```


### Verify multivariate dispersion

Check that multivariate dispersion is ok with the whole data. 
```{r}
Disp_Dens=betadisper(total.d,total.table$Density_code) # significant difference in dispersion p=0.01
anova(Disp_Dens) #0.06326
permutest(Disp_Dens) # 0.054 --- ok
TukeyHSD(Disp_Dens) # no difference

Disp_Fire=betadisper(total.d,total.table$PrePost) # significant difference in dispersion p=0.01
anova(Disp_Fire) #0.0001
permutest(Disp_Fire) # 0.001

Disp_Grid=betadisper(total.d,total.table$GRID_ID) # significant difference in dispersion p=0.01
anova(Disp_Grid) #0.005047
permutest(Disp_Grid) # 0.009

DispMoss3=betadisper(total.d,interaction(total.table$GRID_ID,total.table$PrePost)) # significant diffenrence in dispersion p=0.01
anova(DispMoss3) # <0.0001
permutest(DispMoss3) # 0.001
TukeyHSD(DispMoss3) # everything except the two post fire grids differ

DispMoss4=betadisper(total.d,interaction(total.table$GRID_ID,total.table$PrePost, total.table$Density_code)) # significant diffenrence in dispersion p=0.01
anova(DispMoss4) # 0.0007534
permutest(DispMoss4) # 0.016

TukeyHSD(DispMoss4) # Only one pair differs from one another though. 
# no difference

```

There are issues with multivariate dispersion that are exacerbated in this frequency dataset compared to the cover dataset: 
In Jean et al. 2017: "Multivariate dispersion differed between the different successional stages but not among forest types. This is a well-recognized issue with multivariate analyses, which can lead to confusion between within-group variation (dispersion) and mean values of the groups. PerMANOVA is thought to be less sensitive to this issue than some of its alternatives (ANOSIM or MRPP) (Anderson 2001)."

In our data, the main issue seems to come from using the two grids together. One approach to verify that this is not confounding our findings is the also run separate analyses for each grid. - **See what you get and decide what to include where**

Check multivariate within each grid for PrePost, density and interaction


### Run permanovas
Run the perMANOVA to see whether there is a difference in species composition between pre- and post-fire and between spruce density in thre categories. I first ran the model with the interaction, but since it was not significant the interaction was dropped from the final model. The strata argument indicates the random gripoint (site) effect. 
```{r}
# All sites and grids
set.seed(5)
fit1<- adonis(formula = total.d ~ PrePost*Density_code,
                                data = total.table,
                                strata=total.table$Grid_plot, 
                                perm = 1000)
fit1 # No interaction

# Remove non significant interaction 
set.seed(5)
fit2<- adonis(formula = total.d ~ PrePost+Density_code,
                                data = total.table,
                                strata=total.table$Grid_plot, 
                                perm = 1000)
fit2

pairwise.adonis(x=spe,factors=total.table$Density_code, sim.function='vegdist',sim.method='bray',p.adjust.m='holm')
# Difference between low and high. 
```


##### All grids
Run the NMDS with two dimensions for all sites and grids. 
```{r run the nmds}
set.seed(7)
tp.nmds <- metaMDS(total.d, k = 2, trymax = 100)
stressplot(tp.nmds)
plot(tp.nmds, type="text")
tp.nmds$stress #0.15 not too bad
```

```{r}
tp.scores <- as.data.frame(scores(tp.nmds))
tp.scores$gridpoint <- denali$Grid_plot
tp.scores$Treatment=denali$PrePost
tp.scores$grid=denali$GRID_ID
tp.scores$Density=total.table$Density_code
tp.scores$inter=interaction(tp.scores$Density, tp.scores$Treatment)
levels(tp.scores$inter)
tp.scores$inter <- factor(tp.scores$inter, levels = c("low.pre", "mid.pre", "high.pre","low.post","mid.post","high.post"))
```

```{r}
# Ellipses for interaction
NMDS = data.frame(MDS1 = tp.nmds$points[,1], MDS2 = tp.nmds$points[,2],group=tp.scores$inter)
NMDS.mean=aggregate(NMDS[,1:2],list(group=NMDS$group),mean)

  veganCovEllipse<-function (cov, center = c(0, 0), scale = 1, npoints = 100) 
  {
    theta <- (0:npoints) * 2 * pi/npoints
    Circle <- cbind(cos(theta), sin(theta))
    t(center + scale * t(Circle %*% chol(cov)))
  }

  df_ell <- data.frame()
  for(g in levels(NMDS$group)){
    df_ell <- rbind(df_ell, cbind(as.data.frame(with(NMDS[NMDS$group==g,],
                    veganCovEllipse(cov.wt(cbind(MDS1,MDS2),wt=rep(1/length(MDS1),length(MDS1)))$cov,center=c(mean(MDS1),mean(MDS2)))))
                    ,group=g))
  }

# Ellipses for fire
NMDS.fire = data.frame(MDS1 = tp.nmds$points[,1], MDS2 = tp.nmds$points[,2],group=tp.scores$Treatment)
NMDS.mean.fire=aggregate(NMDS[,1:2],list(group=NMDS$group),mean)

  df_ell_fire <- data.frame()
  for(g in levels(NMDS.fire$group)){
    df_ell_fire <- rbind(df_ell_fire, cbind(as.data.frame(with(NMDS.fire[NMDS.fire$group==g,],
                    veganCovEllipse(cov.wt(cbind(MDS1,MDS2),wt=rep(1/length(MDS1),length(MDS1)))$cov,center=c(mean(MDS1),mean(MDS2)))))
                    ,group=g))
  }

# Ellipses for density  
NMDS.dens = data.frame(MDS1 = tp.nmds$points[,1], MDS2 = tp.nmds$points[,2],group=tp.scores$Density)
NMDS.mean.dens=aggregate(NMDS.dens[,1:2],list(group=NMDS$group),mean)

  df_ell_dens <- data.frame()
  for(g in levels(NMDS.dens$group)){
    df_ell_dens <- rbind(df_ell_dens, cbind(as.data.frame(with(NMDS.dens[NMDS.dens$group==g,],
                    veganCovEllipse(cov.wt(cbind(MDS1,MDS2),wt=rep(1/length(MDS1),length(MDS1)))$cov,center=c(mean(MDS1),mean(MDS2)))))
                    ,group=g))
  }
  
  
  
  # Fix the oder of factors so that the color order is the same

levels(tp.scores$inter)
df_ell$group=(as.factor(df_ell$group))
levels(df_ell$group)
df_ell$group <- factor(df_ell$group, levels = c(levels(tp.scores$inter)))
levels(df_ell$group)

lightgreen="#A1D99B"
mediumgreen="#238B45"
darkgreen="#00441B"

p1 <- ggplot() +
  #geom_path(data=df_ell_fire, aes(x=MDS1, y=MDS2, color=group), size=0.75, linetype=1)+
  geom_path(data = tp.scores, aes(x = NMDS1, y = NMDS2, group=gridpoint), col="grey")+
  geom_point(data = tp.scores, aes(x = NMDS1, y = NMDS2, shape=grid, fill=inter), size=3., col="black")+
  #geom_path(data=df_ell, aes(x=MDS1, y=MDS2, color=group), size=0.75, linetype=1)+
  geom_path(data=df_ell_dens, aes(x=MDS1, y=MDS2, color=group), size=0.75, linetype=1)+
  scale_shape_manual(values=c(21,24))+
  scale_fill_manual(values=c(lightgreen, mediumgreen, darkgreen, "gray86","gray44","gray10"))+
  #scale_color_manual(values=c(lightgreen, mediumgreen, darkgreen, "gray86","gray44","gray10"))+
  scale_color_manual(values=c(darkgreen, lightgreen, mediumgreen))+
  theme_bw()+ 
  coord_fixed()+  ## need aspect ratio of 1!
  theme(legend.position = 'right', 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  annotate("text", x = -0.4, y = 0.65, label = "a)")
p1
```

Species envfit since scores are not available anymore
```{r}
vec.sp <- envfit(tp.scores[,c(1,2)], spe, perm=999, p.max=0.05)
scores(vec.sp, display = "vectors")
vec.sp.df <- as.data.frame(scores(vec.sp, display = "vectors"))
vec.sp.df$species <- rownames(vec.sp.df)

spp.scrs <- as.data.frame(scores(vec.sp, display = "vectors"))
spp.scrs <- cbind(spp.scrs, Species = rownames(spp.scrs))

## Select only significant species
ef<-envfit(tp.scores[,c(1,2)], spe, perm=999) #### Select the good columns !!!!!!
ef.df<-as.data.frame(ef$vectors$arrows*sqrt(ef$vectors$r))
ef.df$species<-rownames(ef.df)

#only significant pvalues
#shortcutting ef$vectors
A <- as.list(ef$vectors)
#creating the dataframe
pvals<-as.data.frame(A$pvals)
arrows<-as.data.frame(A$arrows*sqrt(A$r))
C<-cbind(spp.scrs, pvals) ### Using the coordinates obtained from Gavin Simpson's answer
#subset
Cred<-subset(C,pvals<0.002) # Too many!
Cred <- cbind(Cred, Species = rownames(Cred))
dim(Cred)
```

ggplot it
```{r NMDS plots}
p1 <- ggplot() +
  #geom_path(data=df_ell_fire, aes(x=MDS1, y=MDS2, color=group), size=0.75, linetype=1)+
  geom_path(data = tp.scores, aes(x = NMDS1, y = NMDS2, group=gridpoint), col="grey")+
  geom_point(data = tp.scores, aes(x = NMDS1, y = NMDS2, shape=grid, fill=inter), size=3., col="black")+
  #geom_path(data=df_ell, aes(x=MDS1, y=MDS2, color=group), size=0.75, linetype=1)+
  geom_path(data=df_ell_dens, aes(x=MDS1, y=MDS2, color=group), size=0.75, linetype=1)+
  scale_shape_manual(values=c(21,24))+
  scale_fill_manual(values=c(lightgreen, mediumgreen, darkgreen, "gray86","gray44","gray10"))+
  #scale_color_manual(values=c(lightgreen, mediumgreen, darkgreen, "gray86","gray44","gray10"))+
  scale_color_manual(values=c(darkgreen, lightgreen, mediumgreen))+
  theme_bw()+ 
  coord_fixed()+  ## need aspect ratio of 1!
  theme(legend.position = 'none', 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  annotate("text", x = -0.4, y = 0.65, label = "a)")
p1

# Half the length of the species arrows to make the plots more compact
p2<- ggplot() +
    geom_path(data = tp.scores, aes(x = NMDS1, y = NMDS2, group=gridpoint), col="grey")+
  geom_segment(data = Cred[,-5],
               aes(x = 0, xend = NMDS1/2, y = 0, yend = NMDS2/2),
               arrow = arrow(length = unit(0.25, "cm")), colour = "gray32")+ # remove the duplicated column
  geom_point(data = tp.scores, aes(x = NMDS1, y = NMDS2,  fill=interaction(Density, Treatment) ,shape=grid), size=3)+#»
  scale_shape_manual(values=c(21,24))+
  scale_fill_manual(values=c( "gray86","gray44","gray10", lightgreen, mediumgreen, darkgreen))+
  scale_color_manual(values=c("gray10", "#00441B" ,"gray86", "#A1D99B","gray44","#238B45" ))+
  theme_bw()+ 
  coord_fixed()+  ## need aspect ratio of 1!
  geom_text_repel(data = Cred[,-5], aes(x = NMDS1/2, y = NMDS2/2, label = Species),size = 3 ,max.overlaps = 75)+
  theme_bw()+ 
  theme(legend.position = 'none', 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  labs(x="NMDS1", y="NMDS2")+
  annotate("text", x = -0.5, y = 0.52, label = "b)")
p2


#Make a plot with p1 showing sites and link, p2 showing species
ggarrange(p1, p2, ncol = 2)

```